---
title: "Regression"
author: "Anna Yeaton"
date: "Fall 2019"
output:
  html_document:
    df_print: paged
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = T)
```

# Lab Section

In this lab, we will go over regression. We will be using the caret package in R. https://topepo.github.io/caret/train-models-by-tag.html

# Perfomance Metrics 

## Residual 

Deviation of the observed value to the estimated value (sample mean)
$$residual=y_i - \hat{y_i}$$
where $\hat{y_i}$ is the estimated value

## Mean Squared Error (MSE)

$$MSE=\frac{1}{n}\sum_{i=1}^{n}(y_i - \hat{y_i})^2$$

## Root Mean Squared Error (RMSE)
Same units as original data.

$$RMSE=\sqrt{MSE}$$

## L2 regularization : Ridge regression. Regularize by adding the sum of the coefficients, squared, to the function. 

$$Ridge Regression=\sum_{i=1}^{n}(y_i - w_0 - \sum_{j=1}^{p}w_jx_{ij})^2 + \lambda\sum_{j=1}^p(w_j)^2$$

## L1 regularization : Lasso Regression. Regularize by adding the sum of the absolute value of the coefficients to the model. Coefficient estimates may be pushed to zero -- Lasso can perform variable selection

$$Lasso Regression=\sum_{i=1}^{n}(y_i - w_0 - \sum_{j=1}^{p}w_jx_{ij})^2 + \lambda\sum_{j=1}^p|w_j|$$


\newpage

### The broad steps of Machine learning in R. 

1. Split the data into training and test. Set test aside. 

2. Fit a good model to the training data. 

3. See how your model did on the training data.

4. Test how your model performs on the test data. 

# Regression

```{r, include=FALSE}
library(caret)
library(MASS)
library(ggplot2)
library(dplyr)
library(ggfortify)


#Mauna Loa CO2 concentrations
data(airquality)
```


1. Split data into training and test set (75% in train set, 25% in test set)

```{r}
#train_regression = 
#test_regression = 
```


### Linear Regression

* Assumes a linear relationship. 
* Independent variables should not be correlated (no mulitcollinearity)
* The number of observations should be greater than the number of independent variables.


$$RSS=\sum(y_i - \hat{y_i})^2$$
We will predict the response of the Temperature based on Wind. 

This is the data we will fit a linear model to. 
```{r}
# ggplot(data = train_regression) +
#   geom_point(aes(x=Wind, y=Temp)) +
#   theme_bw()
```

2. Create and fit a linear model to predict Temperature from Wind using the training set

```{r}
#help(train)

#linear_regression <- train( ~ , data= , method = "lm")
```


3. Vizualize how your model performed on the train data by plotting the regression line on top of the train data points. 
```{r}

```


4. Explore how the model performs on the test data. For Linear Regression:

* The residuals should be close to zero.
* There should be equal variance around the regression line (homoscedasticity).
* Residuals should be normally distributed.
* Independent variables and residuals should not be correlated.

4 a) See how the model performs on the test data
```{r}
#help(predict)
#linear_predict <- predict(, newdata=)
```

4 b) Look at the residuals. Are they close to zero?
```{r}
#look at the median residual value. Close to zero is best
#help(summary)
```


4 c) Plot predicted temperature vs observed temperature. A strong model should show a strong correlation
```{r}

```

4 d) Visualize the predicted values in relation to the read data points. Look for homoscedasticity
```{r}
# Extract coefficients from the model

# plot the regression line on the predicted values

# plot the original test values

```

4 e) Residuals should be normally distributed. Plot the density of the residuals
```{r}
#residuals_lin <- residuals(linear_regression)

#ggplot(data=residvpredict) +
#  geom_density(aes(residual))
```


4 f) Independent variables and residuals should not be correlated
```{r}
#cor.test(train_regression$Wind, resid(linear_regression))
```


### Linear Regression with Regularization

5. Create a linear model using L1 or L2 regularization to predict Temperature from Wind and Month variables.  



